---
title: "Quercus Plastome VCF Basic Statistics"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.8
)

library(ggplot2)

stats_dir <- file.path(getwd(), "bcftools_stats")
stats_files <- sort(list.files(stats_dir, pattern = "\\.stats\\.txt$", full.names = TRUE))

clean_vcf_name <- function(path) {
  sub("\\.stats\\.txt$", "", basename(path))
}

read_section <- function(path, tag) {
  lines <- readLines(path, warn = FALSE)
  section_lines <- grep(paste0("^", tag, "\t"), lines, value = TRUE)
  if (!length(section_lines)) {
    return(data.frame())
  }
  fields <- strsplit(section_lines, "\t", fixed = TRUE)
  width <- max(vapply(fields, length, integer(1)))
  mat <- t(vapply(fields, function(x) {
    length(x) <- width
    x
  }, character(width)))
  as.data.frame(mat, stringsAsFactors = FALSE)
}

as_numeric_columns <- function(x, skip = character()) {
  for (nm in setdiff(names(x), skip)) {
    suppressWarnings({
      converted <- as.numeric(x[[nm]])
    })
    if (!all(is.na(converted) & !is.na(x[[nm]]))) {
      x[[nm]] <- converted
    }
  }
  x
}

parse_summary <- function(path) {
  sn <- read_section(path, "SN")
  if (!nrow(sn)) {
    return(data.frame())
  }
  names(sn)[1:4] <- c("section", "id", "metric", "value")
  sn$metric <- sub(":$", "", sn$metric)
  sn$value <- as.numeric(sn$value)
  wide <- as.data.frame(t(setNames(sn$value, sn$metric)), check.names = FALSE)
  wide$vcf <- clean_vcf_name(path)
  wide[, c("vcf", setdiff(names(wide), "vcf"))]
}

parse_tstv <- function(path) {
  x <- read_section(path, "TSTV")
  if (!nrow(x)) {
    return(data.frame())
  }
  names(x)[1:8] <- c(
    "section", "id", "transitions", "transversions", "ts_tv",
    "transitions_first_alt", "transversions_first_alt", "ts_tv_first_alt"
  )
  x$vcf <- clean_vcf_name(path)
  x <- x[, c("vcf", names(x)[!(names(x) %in% c("vcf", "section", "id"))])]
  as_numeric_columns(x, skip = "vcf")
}

parse_af <- function(path) {
  x <- read_section(path, "AF")
  if (!nrow(x)) {
    return(data.frame())
  }
  names(x)[1:10] <- c(
    "section", "id", "allele_frequency", "snps", "transitions",
    "transversions", "indels", "repeat_consistent",
    "repeat_inconsistent", "not_applicable"
  )
  x$vcf <- clean_vcf_name(path)
  x <- x[, c("vcf", names(x)[!(names(x) %in% c("vcf", "section", "id"))])]
  as_numeric_columns(x, skip = "vcf")
}

parse_qual <- function(path) {
  x <- read_section(path, "QUAL")
  if (!nrow(x)) {
    return(data.frame())
  }
  names(x)[1:7] <- c(
    "section", "id", "quality", "snps",
    "transitions_first_alt", "transversions_first_alt", "indels"
  )
  x$vcf <- clean_vcf_name(path)
  x <- x[, c("vcf", names(x)[!(names(x) %in% c("vcf", "section", "id"))])]
  as_numeric_columns(x, skip = "vcf")
}

summary_df <- do.call(rbind, lapply(stats_files, parse_summary))
tstv_df <- do.call(rbind, lapply(stats_files, parse_tstv))
af_df <- do.call(rbind, lapply(stats_files, parse_af))
qual_df <- do.call(rbind, lapply(stats_files, parse_qual))

summary_df$variant_density_per_sample <- with(
  summary_df,
  `number of records` / `number of samples`
)

variant_long <- data.frame(
  vcf = rep(summary_df$vcf, each = 4),
  variant_class = rep(c("SNPs", "Indels", "Multiallelic sites", "Multiallelic SNP sites"), nrow(summary_df)),
  count = c(
    rbind(
      summary_df$`number of SNPs`,
      summary_df$`number of indels`,
      summary_df$`number of multiallelic sites`,
      summary_df$`number of multiallelic SNP sites`
    )
  )
)
```

Generated on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")` from `r length(stats_files)` VCF files using `bcftools stats` outputs in `r stats_dir`.

# Input VCFs

```{r input-files}
knitr::kable(
  data.frame(
    vcf = clean_vcf_name(stats_files),
    stats_file = basename(stats_files),
    stringsAsFactors = FALSE
  ),
  caption = "VCF files included in this report."
)
```

# Summary Counts

```{r summary-table}
summary_table <- summary_df[, c(
  "vcf",
  "number of samples",
  "number of records",
  "number of SNPs",
  "number of indels",
  "number of multiallelic sites",
  "number of multiallelic SNP sites",
  "variant_density_per_sample"
)]
summary_table$variant_density_per_sample <- round(summary_table$variant_density_per_sample, 2)
knitr::kable(summary_table, caption = "Core summary statistics from bcftools stats.")
```

```{r variant-counts-plot}
ggplot(variant_long, aes(x = vcf, y = count, fill = variant_class)) +
  geom_col(position = "dodge", width = 0.72) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = NULL,
    y = "Count",
    fill = "Variant class",
    title = "Variant composition by VCF"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

# Transitions And Transversions

```{r tstv-table}
if (nrow(tstv_df)) {
  knitr::kable(tstv_df, caption = "Transition/transversion statistics.")
} else {
  cat("No TSTV section was present in the bcftools stats outputs.")
}
```

```{r tstv-plot}
if (nrow(tstv_df)) {
  ggplot(tstv_df, aes(x = vcf, y = ts_tv)) +
    geom_col(fill = "#2f6f73", width = 0.65) +
    geom_text(aes(label = round(ts_tv, 2)), vjust = -0.35, size = 4) +
    labs(
      x = NULL,
      y = "Ts/Tv ratio",
      title = "Transition/transversion ratio"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 25, hjust = 1),
      panel.grid.minor = element_blank()
    )
}
```

# Allele Frequency Spectrum

```{r af-plot}
if (nrow(af_df)) {
  af_plot_df <- af_df
  af_plot_df$total_variants <- af_plot_df$snps + af_plot_df$indels
  ggplot(af_plot_df, aes(x = allele_frequency, y = total_variants, color = vcf)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.7, alpha = 0.85) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      x = "Non-reference allele frequency",
      y = "SNPs + indels",
      color = "VCF",
      title = "Allele frequency spectrum"
    ) +
    theme_minimal(base_size = 13) +
    theme(panel.grid.minor = element_blank())
} else {
  cat("No AF section was present in the bcftools stats outputs.")
}
```

# Quality Distribution

```{r qual-plot}
if (nrow(qual_df)) {
  qual_df$variant_count <- qual_df$snps + qual_df$indels
  qual_plot_df <- aggregate(variant_count ~ vcf + quality, data = qual_df, sum)
  ggplot(qual_plot_df, aes(x = quality, y = variant_count, color = vcf)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_y_continuous(labels = scales::comma) +
    labs(
      x = "QUAL",
      y = "Variant count",
      color = "VCF",
      title = "Variant counts by QUAL"
    ) +
    theme_minimal(base_size = 13) +
    theme(panel.grid.minor = element_blank())
} else {
  cat("No QUAL section was present in the bcftools stats outputs.")
}
```

# Notes

- Counts are from `bcftools stats`; rows containing multiple variant types can be counted in more than one class.
- The filtered biallelic VCF contains only SNP rows in this report.
- The allele-frequency and quality plots are shown only for sections present in the corresponding stats files.
